Statistical Inference for inter-arrival times of extreme events in bursty time series

### 2020-01-16 article_springer/article_springer.tex
Update ... submitted version

% !TeX program = pdfLaTeX
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{amsmath}
\usepackage{graphicx}

\usepackage[hyphens]{url} % not crucial - just used below for the URL
\usepackage{hyperref}
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%

%% load any required packages here
\usepackage{amssymb}

\begin{document}

\title{Peaks Over Threshold for Bursty Time Series \thanks{Peter Straka was supported by the Discovery Early Career Research Award
DE160101147 on the Project ``Predicting Extremes when Events Occur in
Bursts'' by the Australian Research Council. Katharina Hees was
supported by the DAAD co-financed by the German Federal Ministry of
Education and Research (BMBF).} }

\author{  Katharina Hees \and  Smarak Nayak \and  Peter Straka* \and  }

    \authorrunning{ K. Hees, S. Nayak \& P. Straka }

\institute{
        Katharina Hees \at
     University of Dortmund \\
     \email{\href{mailto:hees@statistik.uni-dortmund.de}{\nolinkurl{hees@statistik.uni-dortmund.de}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Smarak Nayak \at
     National Australia Bank \\
     \email{\href{mailto:smarak.nayak@nab.com.au}{\nolinkurl{smarak.nayak@nab.com.au}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Peter Straka* \at
     UNSW Sydney \\
     \email{\href{mailto:p.straka@unsw.edu.au}{\nolinkurl{p.straka@unsw.edu.au}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
    }

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

\maketitle

@@

This article generalizes the Peaks Over Threshold (POT) model to the
setting where inter-event times are heavy-tailed. For high thresholds
and infinite-mean waiting times, we show that the times between
threshold crossings are Mittag-Leffler distributed, and thus form a
``fractional Poisson Process'' which generalizes the standard Poisson
Process. We provide graphical means of estimating model parameters and
assessing model fit. Along the way, we apply our inference method to a
real-world bursty time series, and show how the memory of the
Mittag-Leffler distribution affects the predictive distribution for the
time until the next extreme event.
\\
\keywords{
        heavy tails \and
        renewal process \and
        extreme value theory \and
        peaks over threshold \and
    }

\end{abstract}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

\section{Introduction}\label{introduction}

Time series displaying temporally inhomogeneous behaviour have received
strong interest in the recent statistical physics literature (Barabási
2005; J. Oliveira and Barabási 2005; Vasquez et al. 2006; Vazquez et al.
2007; Omi and Shinomoto 2011; Min, Goh, and Vazquez 2011; Karsai et al.
2011; Bagrow and Brockmann 2013). They have been observed in the context
of earthquakes, sunspots, neuronal activity and human communication (see
Karsai et al. 2012; Vajna, Tóth, and Kertész 2013; Mark M Meerschaert
and Stoev 2008 for a list of references). Such time series exhibit high
activity in some `bursty' intervals, which alternate with other, quiet
intervals. Although several mechanisms are plausible explanations for
bursty behaviour (most prominently self-exciting point process by Hawkes
(1971)), there seems to be one salient feature which very typically
indicates the departure from temporal homogeneity: a heavy-tailed
distribution of waiting times (Vasquez et al. 2006; Karsai et al. 2012;
Vajna, Tóth, and Kertész 2013). As we show below in simulations, a
simple renewal process with heavy-tailed waiting times can capture this
type of dynamics. For many systems, the renewal property is appropriate;
a simple test of the absence of correlations in a succession of waiting
times can be undertaken by randomly reshuffling the waiting times
(Karsai et al. 2012).

Often a magnitude, or mark can be assigned to each event in the renewal
process, such as for earthquakes, solar flares or neuron voltages. The
Peaks-Over-Threshold model (POT, see e.g. Coles 2001) applies a
threshold to the magnitudes, and fits a Generalized Pareto distribution
to the threshold exceedances. A commonly made assumption in POT models
is that times between events are either fixed or light-tailed, and this
entails that the threshold crossing times form a Poisson process (Hsing,
Hüsler, and Leadbetter 1988). Then as one increases the threshold and
thus decreases the threshold crossing probability \(p\), the Poisson
process is rarefied, i.e.~its intensity decreases \emph{linearly} with
\(p\) (see e.g. Beirlant et al. 2006).

As will be shown below, in the heavy-tailed waiting time scenario
threshold crossing times form a \emph{fractional Poisson process}
(Laskin 2003; Mark M Meerschaert, Nane, and Vellaisamy 2011), which is a
renewal process with Mittag-Leffler distributed waiting times. The
family of Mittag-Leffler distributions nests the exponential
distribution (Haubold, Mathai, and Saxena 2011), and hence the
fractional Poisson process generalizes the standard Poisson process.
Again as the threshold size increases and the threshold crossing
probability \(p\) decreases, the fractional Poisson process is rarefied:
The scale parameter of the Mittag-Leffler inter-arrival times of
threshold crossing times increases, but \emph{superlinearly}; see the
Theorem below.

Maxima of events which occur according to a renewal process with
heavy-tailed waiting times have been studied under the names
``Continuous Time Random Maxima process'' (CTRM) (Benson, Schumer, and
Meerschaert 2007; Mark M Meerschaert and Stoev 2008; Hees and Scheffler
2016; Hees and Scheffler 2017), ``Max-Renewal process'' (Silvestrov
2002; Silvestrov and Teugels 2004; Basrak and Špoljarić 2015), and
``Shock process'' (Esary and Marshall 1973; Shanthikumar and Sumita
1983; Shanthikumar and Sumita 1984; Shanthikumar and Sumita 1985;
Anderson 1987; Gut and Hüsler 1999). The existing literature focuses on
probabilistic results surrounding these models. In this work, however,
we introduce a method of inference for this type of model, which is
seemingly not available in the literature.

We review the marked renewal process in Section 2, and derive a scaling
limit theorem for inter-exceedance times in Section 3. We give a
statistical procedure to estimate model parameters via stability plots
in Section 5, but to set the stage we first discuss inference for the
Mittag-Leffler distribution in Section 4. (A simulation study of the
effectiveness of our statistical procedure is given in the appendix.)
Diagnostic plots for model criticism are discussed in Section 6. In
Section 7, we discuss the memory property of the Mittag-Leffler
distribution, and how it affects the predictive distribution for the
time until the next threshold crossing event. Section 8 concludes. For
all statistical computations we have used R (R Core Team 2018). All code
and data used for the analysis in this article has been organized into
an R package \texttt{CTRE} (\url{https://github.com/UNSW-MATH/CTRE}).
The source code for the figures generated in this manuscript is
available online at \url{https://github.com/UNSW-MATH/bursty-POT}.

\section{Continuous Time Random Exceedances
(CTRE)}\label{continuous-time-random-exceedances-ctre}

@@

\begin{description}
\item[\textbf{Definition (MRP):}]
Let \((W,J), (W_1, J_1), (W_2, J_2), \ldots\) be i.i.d. pairs of random
variables, where the \(W_k > 0\) are interpreted as the \emph{waiting
times} and \(J_k \in [x_L, x_R]\) as the \emph{event magnitudes}
(\(x_L \in [-\infty, +\infty), x_R \in (-\infty, +\infty]\)). If \(W\)
and \(J\) are independent, the Marked Renewal Process is said to be
\emph{uncoupled}. \qed
\end{description}

Note that the \(k\)-th magnitude \(J_k\) occurs at time
\(T_k = W_1 + \ldots + W_k\). Based on an MRP, we define the Continuous
Time Random Exceedance model (CTRE) as follows:

\begin{description}
\item[\textbf{Definition (CTRE):}]
Given a threshold \(\ell \in (x_L, x_R)\), consider the stopping time
\[\tau(\ell) := \min\{k: J_k > \ell\},\quad \ell \in (x_L, x_R).\]
Define the pair of random variables \((X(\ell), T(\ell))\) via
\[X(\ell) = J_{\tau(\ell)} - \ell, \quad 
T(\ell) = \sum_{k=1}^{\tau(\ell)} W_k.\] By restarting the MRP at
\(\tau(\ell)\), inductively define the two i.i.d. sequences
\(T(\ell,n)\) and \(X(\ell, n)\), \(n \in \mathbb N\), called the
``interarrival times'' and the ``exceedances'', respectively. The pair
sequence \((T(\ell, n), W(\ell, n))_{n \in \mathbb N}\) is called a
Continuous Time Random Exceedance model (CTRE). If the underlying MRP is
uncoupled, then the CTRE is also called uncoupled. \qed
\end{description}

\begin{figure}
\includegraphics[width=\textwidth]{article_springer_files/figure-latex/thresholdedBursty-1} \caption{\label{fig:thresholdedBursty}Exceedances (red) and Times until Exceedance (durations between blue crosses) for a given threshold $\ell$ (dashed line).}\label{fig:thresholdedBursty}
\end{figure}

In this article, we restrict ourselves to the uncoupled case, where the
two sequences \(X(\ell, n)_{n \in \mathbb N}\) and
\(T(\ell, n)_{n \in \mathbb N}\) are independent.\footnote{To see why,
  note that \(X(\ell)\) is, in distribution, simply equal to
  \(J | J > \ell\), independent of any waiting time \(W_k\).} Figure
\ref{fig:thresholdedBursty} shows a simulated dataset in the top panel,
where \(W\) has a stable distribution with tail parameter \(\beta =\)
0.8 (and skewness \(1\) and location \(0\)), and where \(J\) is from a
standard Gumbel distribution. In the bottom panel, we plot a time series
of solar flare intensities derived from a NASA dataset (Dennis et al.
1991)\footnote{The ``complete Hard X Ray Burst Spectrometer event list''
  is a comprehensive reference for all measurements of the Hard X Ray
  Burst Spectrometer on NASA's Solar Maximum Mission from the time of
  launch on Feb 14, 1980 to the end of the mission in Dec 1989. 12,776
  events were detected, with the ``vast majority being solar flares''.
  The list includes the start time, peak time, duration, and peak rate
  of each event. We have used ``start time'' as the variable for event
  times, and ``peak rate'' as the variable for event magnitudes.}.
Clearly, the simulated data exhibit long intervals \emph{without any}
events, whereas in the real-world dataset events appear continuously.
The threshold exceedances, however, appear to have similar statistical
behaviour in both models. Observations below a threshold are commonly
discarded in Extreme Value Theory (POT approach); likewise, the CTRE
model interprets these observations as noise and filters them out.

\section{Scaling limit of Exceedance Times}\label{sec:scaling}

@@

\begin{description}
\tightlist
\item[\textbf{Theorem:}]
Let the waiting times \(J_k\) be in the domain of attraction of a
positively skewed sum-stable law with stability parameter
\(0 < \beta < 1\); more precisely,

\begin{align} \label{eq:stability}
\frac{W_1 + \ldots + W_n}{b(n)} \overset{d}{\longrightarrow} D, 
\quad n \to \infty
\end{align}

for a function \(b(n)\) which is regularly varying at \(\infty\) with
parameter \(1/\beta\), and where
\(\mathbf E[\exp(-sD)] = \exp(-s^\beta)\). Write
\(p := \mathbf P(J > \ell)\). Then the weak convergence \[
\frac{T(\ell)} {b(1/p)} \to W_\beta \quad \text{ as } \quad \ell \uparrow x_R
\] holds, where the Mittag-Leffler random variable \(W_\beta\) is
defined on the positive real numbers via \[
\mathbf E[\exp(-sW_\beta)] = \frac{1}{1+s^\beta}.
\] \qed
\end{description}

\noindent For a scale parameter \(\sigma > 0\), we write
\({\rm ML}(\beta, \sigma)\) for the distribution of \(\sigma W_\beta\).
The Mittag-Leffler distribution with parameter \(\beta \in (0,1]\) is a
heavy-tailed positive distribution for \(\beta < 1\), with infinite
mean. However, as \(\beta \uparrow 1\), \({\rm ML}(\beta, \sigma)\)
converges weakly to the exponential distribution \({\rm Exp}(\sigma)\).
This means that although its moments are all infinite, the
Mittag-Leffler distribution may (if \(\beta\) is close to 1) be
indistinguishable from the exponential distribution, for the purposes of
applied statistics. For a detailed reference on the Mittag-Leffler
distribution, see e.g. Haubold, Mathai, and Saxena (2011), and for
algorithms, see e.g.~the R package \texttt{MittagLeffleR} (Gill and
Straka 2017).

\paragraph{\texorpdfstring{\emph{Proof of
Theorem:}}{Proof of Theorem:}}\label{proof-of-theorem}

We interpret the threshold crossing time \(T(\ell)\) as the hitting time
of the underlying CTRM (Continuous Time Random Maxima) or ``max-renewal
process'', and then utilize a result by Mark M Meerschaert and Stoev
(2008). The running maximum process is defined as \[
M(c) := J_1 \vee \ldots \vee J_{\lfloor c \rfloor},
\] and since we assume that the \(J_k\) have a continuous distribution,
there exist norming functions \(a(c)\) and \(d(c)\) such that \[
\mathbf P\left[ \frac{M(c) - d(c)}{a(c)} \le \ell^* \right] 
\longrightarrow F(\ell^*), \quad t \to \infty
\] where \(F\) is a generalized extreme value distribution, and
\(\ell^*\) is any value from the support of \(F\). The CTRM process is
then defined via \[
V(t) = M(N(t)), \quad t \ge 0
\] where \(N(t)\) is the renewal process associated with the waiting
times \(W_k\): \[
N(t) = \max\{n: W_1 + \ldots + W_n \le t\}.
\] Now a key observation is that \[
T(\ell) = \inf\{t: V(t) > \ell\}, 
\] and that \[
T(\ell) > t \quad \text{ if and only if } \quad V(t) \le \ell.
\] By (Theorem 3.1, Mark M Meerschaert and Stoev 2008), we have the
stochastic process convergence \[
\frac{V(ct) - d(\tilde b(c))}{a(\tilde b(c))} 
\stackrel{d}{\longrightarrow} Y(t), \quad t > 0.
\] where \(Y(t)\) is a time-changed (``subordinated'') extremal process,
and where \(\tilde b(c)\) is a regularly varying norming function which
is \emph{inverse} to \(b(c)\), in the sense that
\(b(\tilde b(c)) \sim c \sim \tilde b(b(c))\).

Without loss of generality, we choose \(\ell^*\) such that
\(F(\ell^*) = 1/e\), and let
\(\ell = a(\tilde b(c)) \ell^* + d(\tilde b(c))\). We may then calculate
\[
\mathbf P\left[ \frac{T(\ell)}{b(1/p)} > t \right]
= \mathbf P[T(\ell) > b(1/p) t]
= \mathbf P[V(ct) \le \ell]
\] where we have substituted \(c = b(1/p)\). Moreover \[
\mathbf P[V(ct) \le \ell]
= \mathbf P\left[ \frac{V(ct) - d(\tilde b(c))}{a(\tilde b(c))} 
\le \frac{\ell - d(\tilde b(c))}{a(\tilde b(c))} \right]
\longrightarrow \mathbf P\left[ Y(t) \le \ell^* \right]
\] Defining the hitting time of level \(\ell^*\) by \(Y(t)\) as
\(\xi_{\ell^*} := \inf\{t: Y(t) > \ell^*\}\), we then have \[
P\left[ Y(t) \le \ell^* \right] = \mathbf P[\xi_{\ell^*} > t] 
= \mathbf P[(-\log F(\ell^*))^{-1/\beta} X^{1/\beta} D > t]
\] by (Proposition 4.2, Mark M Meerschaert and Stoev 2008), where \(X\)
is an exponential random variable with mean \(1\). Using (Theorem 19.1,
Haubold, Mathai, and Saxena 2011), we see that
\(X^{1/\beta} D \sim {\rm ML}(\beta, 1)\), concluding the proof. \qed

\begin{description}
\tightlist
\item[\textbf{Remark:}]
If \(\beta = 1\), the result of the Theorem above is standard, see e.g.
Equation (2.2) in Gut and Hüsler (1999).
\end{description}

\section{Inference for the Mittag-Leffler
distribution}\label{inference-for-the-mittag-leffler-distribution}

\begin{figure}
\includegraphics[width=\textwidth]{article_springer_files/figure-latex/QQ-plots-1} \caption{Pareto vs. Mittag-Leffler distribution, both with same tail exponent 0.8. Top row: solar flare magnitudes data, fitting nicely with a Pareto distribution. Bottom row: simulated Mittag-Leffler data. Left column: static QQ-Estimator plots, assessing a fit against the Pareto distribution. Centre column: dynamic QQ-Estimator plots. Right column: a usual QQ-Plot based on Mittag-Leffler quantiles.\label{fig:QQ-plots}}\label{fig:QQ-plots}
\end{figure}

>>>>

\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}

\usepackage{lineno} % add
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Statistical Inference for inter-arrival times of extreme events in bursty time series},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{5}
% Pandoc toggle for numbering sections (defaults to be off)
% Pandoc header

\begin{document}
\begin{frontmatter}

  \title{Statistical Inference for inter-arrival times of extreme events in
bursty time series}
    \author[a]{Katharina Hees}
   \ead{hees@statistik.tu-dortmund.de}

    \author[b]{Smarak Nayak}
   \ead{smarak.nayak@nab.com.au}

    \author[c]{Peter Straka}
   \ead{p.straka@unsw.edu.au}

      \address[a]{Department of Statistics, TU Dortmund University, Dortmund, Germany}
    \address[b]{National Australia Bank, Melbourne, Australia}
    \address[c]{School of Mathematics and Statistics, UNSW, Sydney, Australia}

@@

  This article proposes a new model for the \emph{inter-exceedance times}
  of events above high thresholds; the threshold exceedances itself are
  modeled via the standard Peaks-Over-Threshold (POT) method. For high
  thresholds and infinite-mean waiting times, we show that the times
  between threshold crossings are Mittag-Leffler distributed, and thus
  form a ``fractional Poisson Process'' which generalizes the standard
  Poisson Process of threshold exceedances. We provide graphical means of
  estimating model parameters and assessing model fit. Along the way, we
  apply our inference method to a real-world bursty time series, and show
  how the memory of the Mittag-Leffler distribution affects the predictive
  distribution for the time until the next extreme event.
  \end{abstract}
   \begin{keyword} heavy tails renewal process extreme value theory peaks over threshold\end{keyword}
 \end{frontmatter}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Time series displaying temporally inhomogeneous behaviour in terms of
the occurrence of events have received strong interest in the recent
statistical physics literature (Barabási 2005; Oliveira and Barabási
2005; Vasquez et al. 2006; Vazquez et al. 2007; Omi and Shinomoto 2011;
Min, Goh, and Vazquez 2011; Karsai et al. 2011; Bagrow and Brockmann
2013). They have been observed in the context of earthquakes, sunspots,
neuronal activity and human communication (see Karsai et al. 2012;
Vajna, Tóth, and Kertész 2013; Meerschaert and Stoev 2008 for a list of
references). Such time series exhibit high activity in some `bursty'
intervals, which alternate with other, quiet intervals. Although several
mechanisms are plausible explanations for bursty behaviour (most
prominently self-exciting point process by Hawkes (1971)), there seems
to be one salient feature which very typically indicates the departure
from temporal homogeneity: a heavy-tailed distribution of waiting times
(Vasquez et al. 2006; Karsai et al. 2012; Vajna, Tóth, and Kertész
2013). As we show below in simulations, a simple renewal process with
heavy-tailed waiting times can capture this type of dynamics. For many
systems, the renewal property is appropriate; a simple test of the
absence of correlations in a succession of waiting times can be
undertaken by randomly reshuffling the waiting times (Karsai et al.
2012).

Often a magnitude, or mark can be assigned to each event in the renewal
process, such as for earthquakes, solar flares or neuron voltages. The
Peaks-Over-Threshold model (POT, see e.g.~Coles 2001) applies a
threshold to the magnitudes, and fits a Generalized Pareto distribution
to the threshold exceedances. A commonly made assumption in POT models
is that times between events are either fixed or light-tailed, and this
entails that the threshold crossing times form a Poisson process (``On
the Exceedance Point Process for a Stationary Sequence'' 1988). Then as
one increases the threshold and thus decreases the threshold crossing
probability \(p\), the Poisson process is rarefied, i.e.~its intensity
decreases \emph{linearly} with \(p\) (see e.g.~Beirlant et al. 2006).

As will be shown below, in the heavy-tailed waiting time scenario
threshold crossing times form a \emph{fractional Poisson process}
(Laskin 2003; Meerschaert, Nane, and Vellaisamy 2011), which is a
renewal process with Mittag-Leffler distributed waiting times. The
family of Mittag-Leffler distributions nests the exponential
distribution (Haubold, Mathai, and Saxena 2011), and hence the
fractional Poisson process generalizes the standard Poisson process.
Again as the threshold size increases and the threshold crossing
probability \(p\) decreases, the fractional Poisson process is rarefied:
The scale parameter of the Mittag-Leffler inter-arrival times of
threshold crossing times increases, but \emph{superlinearly}; see the
Theorem below.

Maxima of events which occur according to a renewal process with
heavy-tailed waiting times have been studied under the names
``Continuous Time Random Maxima process'' (CTRM) (Benson, Schumer, and
Meerschaert 2007; Meerschaert and Stoev 2008; Hees and Scheffler 2018b,
2018a), ``Max-Renewal process'' (Silvestrov 2002; Silvestrov and Teugels
2004; Basrak and Špoljarić 2015), and ``Shock process'' (Esary and
Marshall 1973; Shanthikumar and Sumita 1983, 1984, 1985; Anderson 1987;
Gut and Hüsler 1999). The existing literature focuses on probabilistic
results surrounding these models. In this work, however, we introduce a
method of inference for this type of model, which is seemingly not
available in the literature.

We review the marked renewal process in Section 2, and derive a scaling
limit theorem for inter-exceedance times in Section 3. We give a
statistical procedure to estimate model parameters via stability plots
in Section 5, but to set the stage we first discuss inference for the
Mittag-Leffler distribution in Section 4. A simulation study of the
effectiveness of our statistical procedure is given in Section 6. In
Section 7 we apply our method to a real data set. In Section 8, we
discuss the memory property of the Mittag-Leffler distribution, and how
it affects the predictive distribution for the time until the next
threshold crossing event. Finally we close with a discussion and
conclusion in Section 9. For all statistical computations we have used R
(R Core Team 2018). All code and data used for the analysis in this
article has been organized into an R package \texttt{CTRE}
(\url{https://github.com/strakaps/CTRE}). The source code for the
figures generated in this manuscript is available online at
\url{https://github.com/strakaps/bursty-POT}.

\hypertarget{continuous-time-random-exceedances-ctre}{%
\section{Continuous Time Random Exceedances
(CTRE)}\label{continuous-time-random-exceedances-ctre}}

@@

\begin{description}
\item[\textbf{Definition (MRP):}]
Let \((W,J), (W_1, J_1), (W_2, J_2), \ldots\) be i.i.d. pairs of random
variables, where the \(W_k > 0\) are interpreted as the \emph{waiting
times} and \(J_k \in [x_L, x_R]\) as the \emph{event magnitudes}
(\(x_L \in [-\infty, +\infty), x_R \in (-\infty, +\infty]\)). If \(W\)
and \(J\) are independent, the Marked Renewal Process is said to be
\emph{uncoupled}.
\end{description}

We assume that the \(k\)-th magnitude \(J_k\) occurs at time
\(T_k = W_1 + \ldots + W_k\). Based on an MRP, we define the Continuous
Time Random Exceedance model (CTRE) as follows:

\begin{description}
\item[\textbf{Definition (CTRE):}]
Given a threshold \(\ell \in (x_L, x_R)\), consider the stopping time
\[\tau(\ell) := \min\{k: J_k > \ell\},\quad \ell \in (x_L, x_R).\]
Define the pair of random variables \((X(\ell), T(\ell))\) via
\[X(\ell) = J_{\tau(\ell)} - \ell, \quad 
T(\ell) = \sum_{k=1}^{\tau(\ell)} W_k.\] By restarting the MRP at
\(\tau(\ell)\), inductively define the two i.i.d. sequences
\(T(\ell,n)\) and \(X(\ell, n)\), \(n \in \mathbb N\), called the
``inter-arrival times'' and the ``exceedances'', respectively. The pair
sequence \((T(\ell, n), W(\ell, n))_{n \in \mathbb N}\) is called a
Continuous Time Random Exceedance model (CTRE). If the underlying MRP is
uncoupled, then the CTRE is also called uncoupled.
\end{description}

\begin{figure}

{\centering \includegraphics[width=0.7\linewidth]{article_springer_files/figure-latex/thresholdedBursty-1}

}

\caption{\label{fig:thresholdedBursty} Exceedances (red) and times until Exceedance (durations between blue crosses) for a given threshold $\ell$ (dashed line). Upper picture: Simulated data with stable distributed waiting times. Lower picture: Solar flares during 1982.}\label{fig:thresholdedBursty}
\end{figure}

In this article, we restrict ourselves to the uncoupled case, where
\(W\) and \(J\) are independent. Then the two sequences
\(X(\ell, n)_{n \in \mathbb N}\) and \(T(\ell, n)_{n \in \mathbb N}\)
are independent as well. To see why, note that \(X(\ell)\) is, in
distribution, simply equal to \(J | J > \ell\), independent of any
waiting time \(W_k\).\\
Figure \ref{fig:thresholdedBursty} shows a simulated dataset in the top
panel, where \(W\) has a stable distribution with tail parameter
\(\beta =\) 0.8 (and skewness \(1\) and location \(0\)), and where \(J\)
is from a standard Gumbel distribution. In the bottom panel, we plot a
time series of solar flare intensities derived from a NASA dataset
(Dennis et al. 1991) which we will later examine more closely (see
Section 7). Clearly, the simulated data exhibit long intervals
\emph{without any} events, whereas in the real-world dataset events
appear continuously. The threshold exceedances, however, appear to have
visually similar statistical behaviour in both models. Observations
below a threshold are commonly discarded in Extreme Value Theory (POT
approach); likewise, the CTRE model interprets these observations as
noise and discards them.

\hypertarget{sec:scaling}{%
\section{Scaling limit of Exceedance Times}\label{sec:scaling}}

@@

\begin{description}
\item[\textbf{Theorem:}]
Let the waiting times \(J_k\) be in the domain of attraction of a
positively skewed sum-stable law with stability parameter
\(0 < \beta < 1\); more precisely, \begin{align} \label{eq:stability}
\frac{W_1 + \ldots + W_n}{b(n)} \overset{d}{\longrightarrow} D, 
\quad n \to \infty
\end{align} for a function \(b(n)\) which is regularly varying at
\(\infty\) with parameter \(1/\beta\), and where
\(\mathbf E[\exp(-sD)] = \exp(-s^\beta)\). Write
\(p := \mathbf P(J > \ell)\). Then the weak convergence \[
\frac{T(\ell)} {b(1/p)} \to W_\beta \quad \text{ as } \quad \ell \uparrow x_R
\] holds, where the Mittag-Leffler random variable \(W_\beta\) is
defined on the positive real numbers via \[
\mathbf E[\exp(-sW_\beta)] = \frac{1}{1+s^\beta}.
\]
\end{description}

\emph{Proof of Theorem:} We interpret the threshold crossing time
\(T(\ell)\) as the hitting time of the underlying CTRM (Continuous Time
Random Maxima) or ``max-renewal process'', and then utilize a result by
Meerschaert and Stoev (2008). The running maximum process is defined as
\[
M(c) := J_1 \vee \ldots \vee J_{\lfloor c \rfloor},
\] and since we assume that the \(J_k\) have a continuous distribution,
there exist norming functions \(a(c)\) and \(d(c)\) such that \[
\mathbf P\left[ \frac{M(c) - d(c)}{a(c)} \le \ell^* \right] 
\longrightarrow F(\ell^*), \quad t \to \infty
\] where \(F\) is a generalized extreme value distribution, and
\(\ell^*\) is any value from the support of \(F\). The CTRM process is
then defined via \[
V(t) = M(N(t)), \quad t \ge 0
\] where \(N(t)\) is the renewal process associated with the waiting
times \(W_k\): \[
N(t) = \max\{n: W_1 + \ldots + W_n \le t\}.
\] Now a key observation is that \[
T(\ell) = \inf\{t: V(t) > \ell\}, 
\] and that \[
T(\ell) > t \quad \text{ if and only if } \quad V(t) \le \ell.
\] By (Theorem 3.1, Meerschaert and Stoev 2008), we have the stochastic
process convergence \[
\frac{V(ct) - d(\tilde b(c))}{a(\tilde b(c))} 
\stackrel{d}{\longrightarrow} Y(t), \quad t > 0.
\] where \(Y(t)\) is a time-changed (``subordinated'') extremal process,
and where \(\tilde b(c)\) is a regularly varying norming function which
is \emph{inverse} to \(b(c)\), in the sense that
\(b(\tilde b(c)) \sim c \sim \tilde b(b(c))\).

Without loss of generality, we choose \(\ell^*\) such that
\(F(\ell^*) = 1/e\), and let
\(\ell = a(\tilde b(c)) \ell^* + d(\tilde b(c))\). We may then calculate
\[
\mathbf P\left[ \frac{T(\ell)}{b(1/p)} > t \right]
= \mathbf P[T(\ell) > b(1/p) t]
= \mathbf P[V(ct) \le \ell]
\] where we have substituted \(c = b(1/p)\). Moreover \[
\mathbf P[V(ct) \le \ell]
= \mathbf P\left[ \frac{V(ct) - d(\tilde b(c))}{a(\tilde b(c))} 
\le \frac{\ell - d(\tilde b(c))}{a(\tilde b(c))} \right]
\longrightarrow \mathbf P\left[ Y(t) \le \ell^* \right]
\] Defining the hitting time of level \(\ell^*\) by \(Y(t)\) as
\(\xi_{\ell^*} := \inf\{t: Y(t) > \ell^*\}\), we then have \[
P\left[ Y(t) \le \ell^* \right] = \mathbf P[\xi_{\ell^*} > t] 
= \mathbf P[(-\log F(\ell^*))^{-1/\beta} X^{1/\beta} D > t]
\] by (Proposition 4.2, Meerschaert and Stoev 2008), where \(X\) is an
exponential random variable with mean \(1\). Using (Theorem 19.1,
Haubold, Mathai, and Saxena 2011), we see that
\(X^{1/\beta} D \sim {\rm ML}(\beta, 1)\), concluding the proof. \qed

For a scale parameter \(\sigma > 0\), we write
\({\rm ML}(\beta, \sigma)\) for the distribution of \(\sigma W_\beta\).
The Mittag-Leffler distribution with parameter \(\beta \in (0,1]\) is a
heavy-tailed positive distribution for \(\beta < 1\), with infinite
mean. However, as \(\beta \uparrow 1\), \({\rm ML}(\beta, \sigma)\)
converges weakly to the exponential distribution \({\rm Exp}(\sigma)\)
with mean \(\sigma\). This means that although its moments are all
infinite, the Mittag-Leffler distribution may (if \(\beta\) is close to
1) be indistinguishable from the exponential distribution, for the
purposes of applied statistics. For a detailed reference on the
Mittag-Leffler distribution, see e.g.~ Haubold, Mathai, and Saxena
(2011), and for algorithms, see e.g.~the R package
\texttt{MittagLeffleR} (Gill and Straka 2017).

\begin{description}
\item[\textbf{Remark:}]
If \(\beta = 1\), the result of the Theorem above is standard, see e.g.~
Equation (2.2) in Gut and Hüsler (1999). In Anderson (1987) a similar
result is shown with a different choice of scaling constant.
\end{description}

\hypertarget{sec:ML}{%
\section{Model Choice and inference for the Mittag-Leffler
distribution}\label{sec:ML}}

The classical POT approach is based on the fact that exceedances above a
high threshhold are asymptotically GPD distributed, hence a GPD
distribution is fitted to the exceedances above a high enough
threshhold. Due to the Theorem in the last Section we know that for high
thresholds the times between exceedances are asymptotically
Mittag-Leffler distributed in case of heavy-tailed waiting times between
the observations. Hence, before we talk about inference for the
exceedances in the next section, we discuss inference for Mittag-Leffler
distributions.

Historically, the first method proposed for the estimation of the
Mittag-Leffler distribution parameters was the fractional moment
estimator by Kozubowski (2001). Unlike the first moments, the fractional
moments of order \(p\) for \(p<\beta\) exist and are tractable. One
drawback of this method is that constant priors for the tail parameter
are needed for the calculation of the estimates. Cahoy, Uchaikin, and
Woyczynski (2010) proposed a moment estimator of the log-transformed
data, which does not require any prior. Furthermore, they performed
simulation studies illustrating that the log-moment outperforms the
fractional moment estimator with respect to bias and root mean squared
error (RMSE).

Due to the form of the density function of a Mittag-Leffler distribution
there exists no closed form for the Maximum Likelihood estimator (MLE).
In the R Package \texttt{MittagLeffleR} (Gill and Straka 2017), Maximum
Likelihood estimation is implemented via numerical optimization. The MLE
slightly outperforms the log-moment estimator regarding bias and RMSE
for big enough sample sizes, but is extremely computational intensive.
Figure \ref{fig:MSE} shows that both estimators perform well, even for
small sample sizes.

\begin{figure}

{\centering \includegraphics[width=0.9\linewidth]{article_springer_files/figure-latex/MSE-1}

}

\caption{\label{fig:MSE} MSE for the estimation of tail (left column) and scale (right column) parameters via log-moment estimator and mle of a Mittag-Leffler sample with varying sample size n=25, 100, 200, varying tails on the x-axis and fixed scale equal to one.}\label{fig:MSE}
\end{figure}
